---
title: "House Prices"
author: "Cason Wight"
date: "5/2/2020"
output: html_document
---

The data were obtained from a [kaggle competition](https://www.kaggle.com/c/house-prices-advanced-regression-techniques/data).


##### Loading the Necessary Packages {.tabset .tabset-pills .tabset-fade}

###### R

I will use dplyr and tibble from the tidyverse for data manipulation and ggplot for visualizations.

```{r packages}
library(ggplot2)
library(dplyr)
library(tibble)
```

###### Python

I will use pandas and numpy for data manipulation and matplotlib for visualizations.

```{python pypackages}
import pandas as pd
import numpy as np
import random
import matplotlib.pyplot as plt
```

##### Reading in the Data {.tabset .tabset-pills .tabset-fade}

```{r seed, echo = FALSE}
set.seed(1)
```

```{python seedPy, echo = FALSE}
random.seed(1)
```


###### R 

```{r readData}
train1 <- tibble(read.csv("train.csv")[,-1])
head(train1)
```

###### Python 

```{python readDataPy}
train2 = pd.read_csv("train.csv").iloc[:,1:]
train2.head()
```


